### 2019-09-19 Manuscript/Supplementary_Information/Supplementary_Information.tex
Updates from Overleaf

Yuem Park\textsuperscript{1},
Nicholas L. Swanson-Hysell\textsuperscript{1},
Pierre Maffre\textsuperscript{1},
Francis A. Macdonald\textsuperscript{2},
Eliel A. Anttila\textsuperscript{2},
Yves Godd\'eris\textsuperscript{3}

@@

The silicate weathering component of the GEOCLIM model implements the model of \cite{Gabet2009a} for the development of a chemically weathered profile. We refer to this chemically weathered profile as regolith where the base of the regolith is unweathered bedrock. In the model of \citet{Gabet2009a}, material enters the regolith and leaves either as a solute through chemical weathering as a particle transits through the regolith or as a physically weathered particle once it reaches the top. We use the DynSoil implementation of the \cite{Gabet2009a} model which integrates a climatic dependence on the chemical weathering that occurs within the regolith using the formulation of \cite{West2012a}. The transient time-varying version of this regolith model is described by the following system of three equations which form the basis of DynSoil:

@@

\noindent
Equation \ref{eq:1} is a statement of material conservation, where $h$ is the total height of the regolith (m), $t$ is the time in the model (yr), $P_{r}$ is the regolith production rate (m/yr), and $E_{p}$ is the physical erosion rate (m/yr). Equation \ref{eq:2} describes how the residual fraction of weatherable phases ($x$, unitless) changes as a function of time ($t$, yr) and depth (where $z$ is the height above the base of the regolith; m). $K \tau^{\sigma}$ is the dissolution rate constant, which depends on the local climate (captured by $K$, yr$^{-1-\sigma}$) and the time that a given rock particle has spent in the regolith ($\tau$, yr) to some power $\sigma$ (unitless) which implements a time-dependence. Equation \ref{eq:3} describes how the time that a given rock particle has spent in the regolith changes as time in the model progresses.

@@

\noindent
$P_{0}$ is the `optimal' regolith production rate (m/yr), which is defined to be the regolith production rate when there is no overlying regolith. In Equation \ref{eq:6}, we parameterize this value using the model of \citet{Carretier2014a}, where $k_{rp}$ is a proportionality constant (unitless), $q$ is the runoff (m/yr), $E_{a}$ is the activation energy (J/K/mol), $R$ is the universal gas constant (J/mol), $T$ is the temperature (K), and $T_{0}$ is the reference temperature (K). $f(h)$ is the soil production function (unitless), which describes how regolith production decreases as the thickness of the regolith increases. It takes an exponential form as is commonly applied in the literature (e.g. \citealp{Gabet2009a}). In Equation \ref{eq:7}, we follow \citet{Heimsath1997a}, where $d_{0}$ is a reference regolith thickness (m).

@@

\noindent
$k_{e}$ is a proportionality constant ((m/yr)$^{1-m}$) and $m$ and $n$ are adjustable exponents that are kept as 0.5 and 1 as in \citet{Maffre2018a}. This formulation is similar to the large-scale BQART model of \citet{Syvitski2007a}, but without a temperature dependence and the replacement of elevation with slope, resulting in functional form that more closely resembles the stream power law \citep{Davy2000a}. This form and these exponent values are supported by various compilations (e.g. \citealp{Lague2013a}) although \citet{Lague2013a} also suggests that there is variability in the proportionality constant that is difficult to capture at a global scale.

@@

\noindent
$x(z)$ gives the curve of primary phases varying with height upward from the base of the regolith as shown in Figure \ref{fig:regolith_schematic}. Setting $z$ equal to regolith thickness ($h$) gives $x_s$ which is the proportion of primary phases remaining at the top of the column of regolith. This value controls the overall integrated chemical weathering rate out of the regolith.

@@

To calculate the flux of Ca and Mg associated with chemical weathering, we need to assign the concentration of Ca and Mg ($\chi_{CaMg}$) within the unweathered bedrock (Fig. \ref{fig:regolith_schematic}). Previous implementations of GEOCLIM have used bulk continental crust values across all exposed land. Given that the hypothesis we seek to test involves the varying concentration of cations in different lithologies, we have implemented a lithologically-resolved version of the model.

The spatial distribution of lithologies is sourced from the global lithologic map (GLiM) of \citet{Hartmann2012a}. The raw data takes the form of polygon vectors, where each polygon is assigned one of 16 lithologic categories. We first group these 16 categories into 6 broader categories (metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment; Table \ref{tab:GLiM_to_GEOCLIM}). We then rasterize the polygon vectors to 0.1\degrees $\times$ 0.1\degrees resolution, where each pixel is assigned the lithologic category of the polygon that covers the greatest area in that pixel (i.e. the `mode lithology'; Fig. \ref{fig:world_lithology}). However, to improve the computing time of GEOCLIM, we decrease the resolution of the raster to 0.5\degrees $\times$ 0.5\degrees. To do so, a 3-dimensional 720 $\times$ 360 $\times$ 7 matrix is created, in which the fraction of each 0.5\degrees $\times$ 0.5\degrees pixel covered by each of the 6 lithologic categories (or water/ice) is captured by the extra dimension. In this way, we calculate an area-weighted mean Mg+Ca concentration of the surface in each 0.5\degrees $\times$ 0.5\degrees pixel.

@@

Experimental determinations of the activation energy ($E_a$; Equations \ref{eq:8} and \ref{eq:9}) associated with the weathering of silicate minerals are variable \citep{Brantley2003a}. However, multiple efforts to invert for $E_a$ in basaltic watersheds with varying temperature have yielded values (41.6 $\pm$ 3.2~kJ/mol in \citealp{Li2016a}; 42.3~kJ/mol in \citealp{Dessert2001a}) that are consistent with the lower end of activation energies of Ca+Mg bearing minerals in laboratory experiments such as that for diopside (40.5 $\pm$ 1.7~kJ/mol; \citealp{Knauss1993a}) and for labradorite (42.1~kJ/mol; \citealp{Carroll2005a}. We use the value of 42~kJ/mol in our model runs.

Of the various parameters used in the equations that govern the silicate weathering component of GEOCLIM (see SI), the values of four parameters within the chemical weathering parameterization of \cite{West2012a} are poorly constrained: the proportionality constant that modifies the dependence of dissolution rate on runoff and temperature ($k_{d}$), the proportionality constant that modifies the dependence of dissolution rate on runoff only ($k_{w}$), the power constant that modifies the dependence of dissolution rate on the time that a rock particle has spent in the regolith ($\sigma$), and the proportionality constant that modifies the dependence of regolith production on runoff and temperature ($k_{rp}$)

However, the values of four parameters within the chemical weathering parameterization of \cite{West2012a} are poorly constrained: the proportionality constant that modifies the dependence of dissolution rate on runoff and temperature ($k_{d}$; Equation \ref{eq:9}), the proportionality constant that modifies the dependence of dissolution rate on runoff only ($k_{w}$; Equation \ref{eq:9}), the power constant that modifies the dependence of dissolution rate on the time that a rock particle has spent in the regolith ($\sigma$; Equation \ref{eq:4}), and the proportionality constant that modifies the dependence of regolith production on runoff and temperature ($k_{rp}$; Equation \ref{eq:8}). Furthermore, the Ca+Mg concentrations of metamorphic and siliciclastic sediment grid cells are also difficult to define. We therefore allow these parameters to vary within reasonable bounds during the calibration stage of GEOCLIM (Table \ref{tab:parameter_combinations}).

@@

We then compute spatially-resolved long-term \COtwo consumption (i.e. Ca+Mg fluxes) using present-day runoff, temperature, and slope fields. As described in the main text, we then sum the computed \COtwo consumption over large-scale watersheds that appear in the global compilation of \citet{Gaillardet1999a}, as well as smaller-scale watersheds of the Amazon Basin (HYBAM network) in the compilation of \citet{Moquet2011a, Moquet2016a, Moquet2018a}. 80 watersheds in total are used in this study. We then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these basins:

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/W_vs_r2.pdf}
    \caption{\textbf{A)} Modeled global \COtwo consumption vs. the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of the watersheds. Each point represents model output using one of the 78,000 parameter combinations. \textbf{B)} Same as A, but zoomed to the plotting space with positive $r^{2}$'s. The black line represents the full range of estimates of the present-day global CO$_{2}$ emission rate compiled in \citet{Gerlach2011a}. The black box represents the preferred range of those estimates.}
    \label{fig:W_vs_r2}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/r2_cross_plot.pdf}
    \caption{Modeled vs. observed \COtwo consumption in watersheds around the world. \textbf{A)} Each point represents a single watershed, and the y-value of the point shows the mean value of the 3,339 parameter combinations that produce individual watershed \COtwo consumption fluxes that approximate those estimated in the literature for the present-day. Whiskers extend to the minimum and maximum modeled watershed \COtwo consumption fluxes for the 3,339 parameters combinations. The red dashed line is the 1:1 line, the blue dashed line is the 2:1 line, and the cyan dashed lines are the 5:1 and 10:1 lines. \textbf{B)} Same as A, but only showing the parameter combination that produced that highest $r^{2}$.}
    \label{fig:r2_cross_plot}
\end{figure}

@@

In our implementation, the mean global CO$_{2}$ consumption calculated from these 3,339 parameter combinations is 5.4$\times10^{12}$~mol/yr, with a standard deviation of 0.9$\times10^{12}$~mol/yr. These estimates fall within the range of those independently estimated in the literature for the present-day: the full range of estimates of the present-day global CO$_{2}$ emission rate compiled in \citet{Gerlach2011a} is 3.0--10.0$\times10^{12}$~mol/yr, with the preferred range of those estimates being 3.4--5.9$\times10^{12}$~mol/yr (Fig. \ref{fig:W_vs_r2}B).

@@

The lithologic map used in the ``reduced I\&NG + Sunda Shelf'' scenario (Fig. \ref{fig:ING_fracs}) is identical to that used in the ``reduced I\&NG'' scenario, with the addition of the Sunda Shelf. However, the assignment of lithology, runoff, and slope for pixels on the Sunda Shelf is not trivial, given that the shelf is not currently exposed. As discussed in \textit{Paleoshorelines}, the Sunda Shelf is a flat-lying, relatively stable platform of continental crust. Basement highs between Malaysia/Sumatra and Borneo are typically granitic and bounded by basins dominated by shallow marine clastic fill. However, basement highs are less common in the Gulf of Thailand. We therefore assign all pixels of the Sunda Shelf to be 50\% felsic and 50\% sediment, except in the Gulf of Thailand where pixels are assigned to be 100\% sediment. For the slope, we take the mean slope value (0.0038~m/m) of relatively flat land on eastern Sumatra, and assign this value to all pixels of the Sunda Shelf.  For the runoff, we linearly interpolate along latitude bands between present-day land pixels to obtain runoff values for the presently-submerged Sunda Shelf. This approach preserves large N-S variation in runoff associated with the large-scale Hadley circulation and therefore should be reasonable to first-order, but it does not accurately capture the smaller-scale interactions between land and ocean/atmosphere circulation.

\section*{Paleoshorelines}

Geological maps, stratigraphic data, and previous paleoshoreline compilations were used to calculate the changes in different types of subaerially exposed rocks in I\&NG. Following \citet{Molnar2015a}, we analyzed the area changes of islands that are larger than $\sim$200~km$^{2}$. We also included changes in areas of submerged continental shelves that were previously exposed, like the Sunda Shelf. Larger islands were further divided into regions based upon their position on microplates as defined by \citep{Matthews2016a}. Miocene to present stratigraphic columns were compiled from each region to develop an age model for paleoenvironmental indicators. In general, carbonate strata and thin-bedded siliciclastic strata were assumed to be deposited in a subaqueous marine environment, whereas evidence for exposure and terrestrial siliciclastic deposits were used as indicators of a subaerial paleoenvironment. Following previous paleoshoreline compilations and these environmental indicators, paleoshorelines were outlined in QGIS to calculate areas for the Early Pliocene (5 Ma), Late Miocene (10 Ma), and Middle Miocene (15 Ma). The paleoshoreline reconstructions within the paleogeographic models broadly coincide with sub-Epoch boundaries (i.e. Early Miocene = 23--16~Ma, Middle Miocene = 15.99--11.61~Ma, Late Miocene = 11.6--5.4~Ma, Pliocene =5.39--3.8 Ma) but we recognize that the biostratigraphic resolution is an additional source of uncertainty.

Several new sources for paleoshoreline data have become available since \citet{Molnar2015a}. Particularly, for Sulawesi, we followed the recent stratigraphic compilation and paleoshorelines delineated by \citet{Nugraha2018a}, and in New Guinea, we followed \citet{Gold2017a} and \cite{Harrington2017a}. Although the outlines of our paleoshorelines are significantly different from those used for 5~Ma by \citet{Molnar2015a} our calculated area is comparable (Fig. 1 in main text).

@@

Paleogeographic reconstructions of Peninsular Malaysia suggest that it has been largely exposed over the last 20~Ma \citep{Hall2013b, Hall2002a}. Although the majority of the Sunda Shelf is currently submerged, large portions of this flat-lying, relatively stable platform of continental crust were also emergent throughout the Neogene, and were repeatedly subaerially exposed during the Pleistocene as recently as the Last Glacial Maximum \citep{Hall2002a}. Similarities between the terrestrial biotas of Borneo and mainland Southeast Asia confirm the existence of land bridges on the Sunda Shelf throughout the Miocene \citep{Moss1998a}. During the Eocene through Early Miocene rifting of the South China Sea resulted in subsidence throughout the Sunda Shelf \citep{Morley2013a}. Grabens associated with this subsidence became freshwater rift lakes that later transitioned to partially enclosed inland seas and extensive brackish or saline wetland environments. Extensive palynological analysis suggests widespread swamp environments persisting through the Late Miocene \citep{Morley2013a}. Basement highs (e.g. the Natuna Arch, currently subaerially exposed as the largely granitic island of Natuna off SW Borneo, as well as the granitic Tin Islands south of the Malay Peninsula) are typically bounded by Paleogene to Neogene basins dominated by shallow marine clastic fill \citep{Darmadi2007a}. Although most paleogeographic reconstructions of the region incorporate some degree of exposure of the Sunda Shelf from 20~Ma onwards \citep{Hall2013b, Madon2013a}, \citet{Molnar2015a} omit the Sunda Shelf from their paleogeographic model for 5~Ma.

@@

Separated from the Paleozoic continental core of SW Borneo by the Lupar Line suture zone, the northern portion of Borneo (Sabah) is underlain by ophiolitic basement that extends to Palawan \citep{Hall2008a, Ilao2018a}. The Sarawak Basin and NW Borneo trough offshore Sabah host $\textgreater$10~km-thick Neogene sedimentary sequences, indicative of the extent and duration of exhumation in northern Borneo \citep{Hall2008a}. The Late Oligocene/Early Miocene Sabah Orogeny resulted in the uplift of both Sabah and southern Palawan \citep{Hall2013a}, as well as the obduction of the Palawan ophiolite and Telupid ophiolite of northern Sabah. Under-thrusting of thinned passive-margin continental crust beneath these suprasubduction ophiolites resulted in Early to Middle Miocene exhumation and offshore unconformities \citep{Hall2008a}. Early Miocene sediments in northern Sabah have a provenance from Palawan \citep{Suggate2014a}, indicating that Palawan, like northern Borneo, experienced uplift during the Sabah orogeny. However, initiation of back-arc extension in the Sulu Sea around 19~Ma \citep{Hall2013a} resulted in the subsidence of the eastern Palawan Mountains, eliminating or reducing the role of Palawan as a sediment source for the Borneo trough. The 14~Ma Capoas Granite on Palawan and the 7.5~Ma Kinabalu Granite in Sabah are the products of crustal thinning related to pulses of regional backarc extension \citep{Hall2013a}. Thermochronological data from the Sabah highlands suggest extremely rapid uplift (7~km/m.y.) and exhumation during the latest Miocene and early Pliocene, resulting in the recent formation of Mt. Kinabalu, the tallest peak in Borneo \citep{Cottam2013a}.

@@

The East and Southeast Arms consist predominantly of mafic and ultramafic rock of the East Sulawesi Ophoilite that is exposed for more than 10,000~km$^{2}$ \citep{Monnier1995a}. The East Arm preserves a complete ophiolitic sequence underlain by a metamorphic sole, m\'elange, imbricate continental margin and crystalline basement with a blueschist metamorphic overprint \citep{Silver1983a, Monnier1995a, Parkinson1998a}. Locally, the structural thickness of the ophiolite exceeds 15~km with surface relief towering over 3~km \citep{Kadarusman2004a}. Seventeen igneous K-Ar dates from the East Sulawesi ophiolite range from 93--32~Ma, clustering between 60~Ma and 40~Ma (\citealp{Parkinson1998b} and references therein). K-Ar dates on hornblende from the metamorphic sole yielded cooling ages between 36--23~Ma \citep{Parkinson1998a, Villeneuve2001a}, which dates the initial emplacement of the East Sulawesi ophiolite as the north Sulawesi volcanic arc was underthrusted by the Sula spur (Australian crust) \citep{Silver1983a, Parkinson1998a}; however, despite these relatively old ages of emplacement, it appears that the ophiolites on the East Arm were not substantially subaerially exposed and exhumed until the Miocene.

>>>>

Yuem Park\textsuperscript{1},
Pierre Maffre\textsuperscript{1},
Nicholas L. Swanson-Hysell\textsuperscript{1},
Francis A. Macdonald\textsuperscript{2},
Eliel A. Anttila\textsuperscript{2},
Yves Godd\'eris\textsuperscript{3}

@@

The silicate weathering component of the GEOCLIM model as been improved compare to the previous published version (Goddéris et al, 2017 natgeo). The new module implements the model of \cite{Gabet2009a} for the development of a chemically weathered profile. We refer to this chemically weathered profile as regolith where the base of the regolith is unweathered bedrock. In the model of \citet{Gabet2009a}, material enters the regolith and leaves either as a solute through chemical weathering of the material during its travel from the bedrock towards the surface, oras a physically weathered particle once it reaches the top. We use the DynSoil implementation of the \cite{Gabet2009a} model which integrates a climatic dependence on the chemical weathering that occurs within the regolith using the formulation of \cite{West2012a}. The transient time-varying version of this regolith model is described by the following system of three equations which form the core of the new DynSoil module of GEOCLIM:

@@

\noindent
Equation \ref{eq:1} is a statement of material conservation, where $h$ is the total height of the regolith (m), $t$ is the model time (yr), $P_{r}$ is the regolith production rate (m/yr), and $E_{p}$ is the physical erosion rate (m/yr). Equation \ref{eq:2} describes how the residual fraction of weatherable phases ($x$, unitless) changes as a function of time ($t$, yr) and depth (where $z$ is the height above the base of the regolith; m). $K \tau^{\sigma}$ is the dissolution rate constant, which depends on the local climate (captured by $K$, yr$^{-1-\sigma}$) and the time that a given rock particle has spent in the regolith ($\tau$, yr) to some power $\sigma$ (unitless) which implements a time-dependence. Equation \ref{eq:3} describes how the time that a given rock particle has spent in the regolith changes as time in the model progresses.

@@

\noindent
$P_{0}$ is the `optimal' regolith production rate (m/yr), which is defined to be the regolith production rate when there is no overlying regolith. In Equation \ref{eq:6}, we parameterize this value using the model of \citet{Carretier2014a}, where $k_{rp}$ is a proportionality constant (unitless), $q$ is the runoff (m/yr), $E_{a}$ is the activation energy (J/K/mol), $R$ is the ideal gas constant (J/mol), $T$ is the temperature (K), and $T_{0}$ is the reference temperature (K). $f(h)$ is the soil production function (unitless), which describes how regolith production decreases as the thickness of the regolith increases. It takes an exponential form as is commonly applied in the literature (e.g. \citealp{Gabet2009a}). In Equation \ref{eq:7}, we follow \citet{Heimsath1997a}, where $d_{0}$ is a reference regolith thickness (m).

@@

\noindent
$k_{e}$ is a proportionality constant ((m/yr)$^{1-m}$) and $m$ and $n$ are adjustable exponents that are kept as 0.5 and 1 as in \citet{Maffre2018a}. This formulation is close to the large-scale BQART model of \citet{Syvitski2007a}, but without a temperature dependence and the replacement of elevation with slope, resulting in functional form that more closely resembles the stream power law \citep{Davy2000a}. This formulation and these exponent values are supported by various compilations (e.g. \citealp{Lague2013a}) although \citet{Lague2013a} also suggests that there is a variability in the proportionality constant that is difficult to capture at a global scale.

@@

\noindent
$x(z)$ is the abundance profile of primary phases inside the regolith, varying with height upward from the base of the regolith as shown in Figure \ref{fig:regolith_schematic}. Setting $z$ equal to the regolith thickness ($h$) gives $x_s$ which is the proportion of primary phases remaining at the top of the regolith column. This value controls the overall integrated chemical weathering rate out of the regolith.

@@

We restricted the calculation of the weathering fluxes to the flux of dissolved Mg and Ca originating from continental silicate weathering, as they are the sole contributors to the long term consumption of atmospheric CO2. To calculate this flux associated with chemical weathering, we need to assign the concentration of Ca and Mg ($\chi_{CaMg}$) within the unweathered bedrock (Fig. \ref{fig:regolith_schematic}). Previous implementations of GEOCLIM have used bulk continental crust values across all exposed land, using a concept of diffuse lithology (it was assume that the weathering rates of shield rocks, basaltic rocks, and carbonate rocks was the same for each continental grid element, when this grid element was submitted to the same climatic conditions). Given that the hypothesis we seek to test involves the varying concentration of cations in different lithologies, the diffuse lithology cannot be used anymore, we have implemented a realistic lithologically-resolved version of the model.

The spatial distribution of lithologies is sourced from the global lithologic map (GLiM) of \citet{Hartmann2012a}. The raw data takes the form of polygon vectors, where each polygon is assigned one of 16 lithologic categories. We first group these 16 categories into 6 broader categories (metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment; Table \ref{tab:GLiM_to_GEOCLIM}). We then rasterize the polygon vectors to 0.1\degrees $\times$ 0.1\degrees resolution, where each pixel is assigned the lithologic category of the polygon that covers the greatest area in that pixel (i.e. the `mode lithology'; Fig. \ref{fig:world_lithology}). To improve the computing time of GEOCLIM, we decrease the resolution of the raster to 0.5\degrees $\times$ 0.5\degrees. To do so, a 3-dimensional 720 $\times$ 360 $\times$ 7 matrix is created, in which the fraction of each 0.5\degrees $\times$ 0.5\degrees pixel covered by each of the 6 lithologic categories (or water/ice) is captured by the extra dimension. In this way, we calculate an area-weighted mean Mg+Ca concentration of the surface in each 0.5\degrees $\times$ 0.5\degrees pixel.

@@

Experimental determinations of the activation energy ($E_a$; Equations \ref{eq:8} and \ref{eq:9}) associated with the weathering of silicate minerals are variable \citep{Brantley2003a}. However, multiple efforts to invert for $E_a$ in basaltic watersheds with varying temperature have yielded values (41.6 $\pm$ 3.2~kJ/mol in \citealp{Li2016a}; 42.3~kJ/mol in \citealp{Dessert2001a}) that are consistent with the lower end of activation energies of Ca+Mg bearing minerals in laboratory experiments such as that for diopside (40.5 $\pm$ 1.7~kJ/mol; \citealp{Knauss1993a}) and for labradorite (42.1~kJ/mol; \citealp{Carroll2005a}). We use the value of 42~kJ/mol in our model runs.

Within the equations that govern the silicate weathering component of GEOCLIM, we identify the less constrained parameters: the proportionality constant that modifies the dependence of dissolution rate on runoff and temperature ($k_{d}$; Equation \ref{eq:9}), the proportionality constant that modifies the dependence of dissolution rate on runoff only ($k_{w}$; Equation \ref{eq:9}), the power constant that modifies the dependence of dissolution rate on the time that a rock particle has spent in the regolith ($\sigma$; Equation \ref{eq:4}), and the proportionality constant that modifies the dependence of regolith production on runoff and temperature ($k_{rp}$; Equation \ref{eq:8}). Furthermore, the Ca+Mg concentrations of metamorphic and siliciclastic sediment grid cells are also difficult to define. We therefore allow these parameters to vary within reasonable bounds during the calibration stage of GEOCLIM and test 78,000 unique parameter combinations (Table \ref{tab:parameter_combinations}).

@@

We then compute spatially-resolved long-term \COtwo consumption (i.e. Ca+Mg fluxes) using present-day runoff, temperature, and slope fields. As described in the main text, we then sum the computed \COtwo consumption over large-scale watersheds that appear in the global compilation of \citet{Gaillardet1999a}, as well as smaller-scale watersheds of the Amazon Basin (HYBAM network) in the compilation of \citet{Moquet2011a, Moquet2016a, Moquet2018a}. 80 watersheds in total are used in this study. We calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these basins:

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/W_vs_r2.jpg}
    \caption{\textbf{A)} Modeled global \COtwo consumption vs. the coefficient of determination ($r^{2}$) between modeled and data-constrained \COtwo consumption in each of the watersheds. Each point represents model output using one of the 78,000 parameter combinations. \textbf{B)} Same as A, but zoomed to the plotting space with positive $r^{2}$'s. The black line represents the full range of estimates of the present-day global CO$_{2}$ emission rate compiled in \citet{Gerlach2011a}. The black box represents the preferred range of those estimates.}
    \label{fig:W_vs_r2}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/r2_cross_plot.pdf}
    \caption{Modeled vs. data-constrained \COtwo consumption in watersheds around the world. \textbf{A)} Each point represents a single watershed, and the y-value of the point shows the mean value of the 3,339 parameter combinations that produce individual watershed \COtwo consumption fluxes that approximate those estimated in the literature for the present-day. Whiskers extend to the minimum and maximum modeled watershed \COtwo consumption fluxes for the 3,339 parameters combinations. The red dashed line is the 1:1 line, the blue dashed line is the 2:1 line, and the cyan dashed lines are the 5:1 and 10:1 lines. \textbf{B)} Same as A, but only showing the parameter combination that produced that highest $r^{2}$.}
    \label{fig:r2_cross_plot}
\end{figure}

@@

The mean global CO$_{2}$ consumption calculated from these 3,339 parameter combinations is 5.4$\times10^{12}$~mol/yr, with a standard deviation of 0.9$\times10^{12}$~mol/yr. These estimates fall within the range of those independently estimated in the literature for the present-day: the full range of estimates of the present-day global CO$_{2}$ emission rate compiled in \citet{Gerlach2011a} is 3.0--10.0$\times10^{12}$~mol/yr, with the preferred range of those estimates being 3.4--5.9$\times10^{12}$~mol/yr (Fig. \ref{fig:W_vs_r2}B).

@@

The lithologic map used in the ``reduced I\&NG + Sunda Shelf'' scenario (Fig. \ref{fig:ING_fracs}) is identical to that used in the ``reduced I\&NG'' scenario, with the addition of the Sunda Shelf. However, the assignment of lithology, runoff, and slope for pixels on the Sunda Shelf is not trivial, given that the shelf is not currently exposed. As discussed in \textit{Paleoshorelines}, the Sunda Shelf is a flat-lying, relatively stable platform of continental crust. Basement highs between Malaysia/Sumatra and Borneo are typically granitic and bounded by basins dominated by shallow marine clastic fill. However, basement highs are less common in the Gulf of Thailand. We therefore assign all pixels of the Sunda Shelf to be 50\% felsic and 50\% sediment, except in the Gulf of Thailand where pixels are assigned to be 100\% sediment. For the slope, we take the mean slope value (0.0038~m/m) of relatively flat land on eastern Sumatra, and assign this value to all pixels of the Sunda Shelf.  For the runoff, we linearly interpolate along latitude bands between present-day land pixels to obtain runoff values for the presently-submerged Sunda Shelf. This approach preserves large N-S variation in runoff associated with the large-scale Hadley circulation and therefore should be reasonable to first-order, although it does not accurately capture the smaller-scale interactions between land and ocean/atmosphere circulation.

As with any modeling study, there are limitations with the approach. We consider one of the current limitations to be the climate model component of GEOCLIM that was used in this study (GFDL CM2.0) has a lower spatial resolution (2.0\degrees $\times$ 2.5\degrees) than what can be achieved with the latest climate models and computational resources. While this model is effective in capturing the large-scale climatology, using an updated climate model at higher resolution may result in runoff and temperature fields with different spatial distribution and sensitivities to elevation.

Another limitation associated with using a modern-day climate model, is that in the scenarios with varying shorelines, we hold the slope and lithology to be the same as the present-day in each pixel. This means that while we are changing the amount of emergent land, we are not changing the latitudinal or longitudinal position and are keeping the slope of any emerged pixel the same. Given that similar distributions of lithology and slope are reasonable in the past and that latitudinal translation has been relatively minor, we consider this simplification to be reasonable to first-order. However, implementing high-resolution climate models with both varying shorelines and change in paleogeographic position for these scenarios is valuable future direction, although it would require a better knowledge of the past continental topography, which is always challenging. Additionally, such modeling efforts could give insight into whether associated changes in ocean and atmosphere circulation would have significantly modulated precipitation in I\&NG.  Given that precipitation in this region is significantly driven by large-scale convective upwelling leading to high precipitation over both land and water, the current approach is a reasonable approximation for I\&NG. However, for paleogeographic change that would have very significantly modified precipitation over broad swathes of land such as growth of the Himalaya, this approach of using modern climate model results is not viable.

\section*{Paleoshoreline reconstruction and geological synthesis}

Geological maps, stratigraphic data, and previous paleoshoreline compilations were used to calculate the changes in different types of subaerially exposed rocks in I\&NG. Following \citet{Molnar2015a}, we analyzed the area changes of islands that are larger than $\sim$200~km$^{2}$ (Fig. 1 in main text). We also included changes in areas of submerged continental shelves that were previously exposed, like the Sunda Shelf. Larger islands were further divided into regions based upon their position on microplates as defined by \citep{Matthews2016a}. Miocene to present stratigraphic columns were compiled from each region to develop an age model for paleoenvironmental indicators. In general, carbonate strata and thin-bedded siliciclastic strata were assumed to be deposited in a subaqueous marine environment, whereas evidence for exposure and terrestrial siliciclastic deposits were used as indicators of a subaerial paleoenvironment. Following previous paleoshoreline compilations and these environmental indicators, paleoshorelines were outlined in QGIS to calculate areas for the Early Pliocene (5 Ma), Late Miocene (10 Ma), and Middle Miocene (15 Ma). The paleoshoreline reconstructions within the paleogeographic models broadly coincide with sub-Epoch boundaries (i.e. Early Miocene = 23--16~Ma, Middle Miocene = 15.99--11.61~Ma, Late Miocene = 11.6--5.4~Ma, Pliocene =5.39--3.8 Ma) but we recognize that the biostratigraphic resolution is an additional source of uncertainty.

Several new sources for paleoshoreline data have become available since \citet{Molnar2015a}. Particularly, for Sulawesi, we followed the recent stratigraphic compilation and paleoshorelines delineated by \citet{Nugraha2018a}, and in New Guinea, we followed \citet{Gold2017a} and \cite{Harrington2017a}. Although the outlines of our paleoshorelines are significantly different from those used for 5~Ma by \citet{Molnar2015a} our calculated area is comparable.

@@

Paleogeographic reconstructions of Peninsular Malaysia suggest that it has been largely exposed over the last 20~Ma \citep{Hall2013b, Hall2002a}. Although the majority of the Sunda Shelf is currently submerged, large portions of this flat-lying, relatively stable platform of continental crust were also emergent throughout the Neogene, and were repeatedly subaerially exposed during the Pleistocene as recently as the Last Glacial Maximum \citep{Hall2002a}. Similarities between the terrestrial biotas of Borneo and mainland Southeast Asia confirm the existence of land bridges on the Sunda Shelf throughout the Miocene \citep{Moss1998a}. Eocene through Miocene rifting of the South China Sea resulted in subsidence throughout the Sunda Shelf \citep{Morley2013a}. Grabens associated with this subsidence became freshwater rift lakes that later transitioned to partially enclosed inland seas and extensive brackish or saline wetland environments. Palynological analysis suggests widespread swamp environments persisting through the Late Miocene \citep{Morley2013a}. Basement highs (e.g. the Natuna Arch, currently subaerially exposed as the largely granitic island of Natuna off SW Borneo, as well as the granitic Tin Islands south of the Malay Peninsula) are typically bounded by Paleogene to Neogene basins dominated by shallow marine clastic fill \citep{Darmadi2007a}. Most paleogeographic reconstructions of the region incorporate some degree of exposure of the Sunda Shelf from 20~Ma onwards \citep{Hall2013b, Madon2013a}, although it is omitted in the 5 Ma paleogeographic model of \citet{Molnar2015a}.

@@

Separated from the Paleozoic continental core of SW Borneo by the Lupar Line suture zone, the northern portion of Borneo (Sabah) is underlain by ophiolitic basement that extends to Palawan \citep{Hall2008a, Ilao2018a}. The Sarawak Basin and NW Borneo trough offshore Sabah host $\textgreater$10~km-thick Neogene sedimentary sequences, indicative of the extent and duration of exhumation in northern Borneo \citep{Hall2008a}. The Late Oligocene/Early Miocene Sabah Orogeny resulted in the uplift of both Sabah and southern Palawan \citep{Hall2013a}, as well as the obduction of the Palawan ophiolite and Telupid ophiolite of northern Sabah. Under-thrusting of thinned passive-margin continental crust beneath these suprasubduction ophiolites resulted in Early to Middle Miocene exhumation and offshore unconformities \citep{Hall2008a}. Early Miocene sediments in northern Sabah have a provenance from Palawan \citep{Suggate2014a}, indicating that Palawan, like northern Borneo, experienced uplift during the Sabah orogeny. However, initiation of back-arc extension in the Sulu Sea around 19~Ma \citep{Hall2013a} resulted in the subsidence of the eastern Palawan Mountains, eliminating or reducing the role of Palawan as a sediment source for the Borneo trough. The 14~Ma Capoas Granite on Palawan and the 7.5~Ma Kinabalu Granite in Sabah are interpretted to be the result of crustal thinning related to pulses of regional backarc extension \citep{Hall2013a}. Thermochronological data from the Sabah highlands suggest extremely rapid uplift (7~km/m.y.) and exhumation during the latest Miocene and early Pliocene, resulting in the recent formation of Mt. Kinabalu, the tallest peak in Borneo \citep{Cottam2013a}.

@@

The East and Southeast Arms consist predominantly of mafic and ultramafic rock of the East Sulawesi Ophoilite that is exposed for more than 10,000~km$^{2}$ \citep{Monnier1995a}. The East Arm preserves a complete ophiolitic sequence underlain by a metamorphic sole, m\'elange, imbricate continental margin and crystalline basement with a blueschist metamorphic overprint \citep{Silver1983a, Monnier1995a, Parkinson1998a}. Locally, the structural thickness of the ophiolite exceeds 15~km with surface relief over 3~km \citep{Kadarusman2004a}. Seventeen igneous K-Ar dates from the East Sulawesi ophiolite range from 93--32~Ma, clustering between 60~Ma and 40~Ma (\citealp{Parkinson1998b} and references therein). K-Ar dates on hornblende from the metamorphic sole yielded cooling ages between 36--23~Ma \citep{Parkinson1998a, Villeneuve2001a}, which dates the initial emplacement of the East Sulawesi ophiolite as the north Sulawesi volcanic arc was underthrusted by the Sula spur (Australian crust) \citep{Silver1983a, Parkinson1998a}; however, despite these relatively old ages of emplacement, it appears that the ophiolites on the East Arm were not substantially subaerially exposed and exhumed until the Miocene.
